Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-05-30 doc/manuscript.Rmd
Updated with modifications, edits, and format for journal.

***Text Length***: 35,640 / 50,000 Characters

@@

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Most previous work has relied on studying microbes and viruses independently, thereby reducing them to two separate communities. Such approaches are unable to capture how these microbial communities interact, such as through processes that maintain community stability or allow phage-host populations to co-evolve. We developed and implemented a network-based analytical approach to describe phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages could infect which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate phage-bacteria connectedness across the human body. We found that gut and skin network structures were person-specific and were not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. This study quantified and contrasted the diversity of virome-microbiome networks across the human body and illustrated how environmental factors may influence phage-bacteria interactive dynamics. This work provides a baseline for future studies to better understand system perturbations, such as disease states, through ecological networks.

**Word Count**: 243 / 250

@@

Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Bacterial communities have been associated with disease states, including a range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases, such as inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz] and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered human viromes (virus communities consisting primarily of bacteriophages) also have been  associated with diseases and perturbations, including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], spread of antibiotic resistance [@Modi:2013fia], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. Viruses act in concert with their microbial hosts as a single ecological community [@Haerter:2014ii]. Viruses influence their living microbial host communities through processes including lysis, modulating host gene expression [@Lindell:2005gz, @Tyler:2013fl, @Hargreaves:2014ja], influencing evolutionary processes, such as horizontal gene transfer [@Moon:2015fa, @Modi:2013fi, @Ogg:1981th, @Frost:2005dn] or antagonistic co-evolution [@Koskella:2014ds], and by altering ecosystem processes and elemental stoichiometry [@Jover:2014gq].

Previous human microbiome work has focused on bacterial and viral communities, but have reduced them to two separate communities by studying them independently  [@Norman:2015kb; @Manrique:2016dx; @Ly:2014ew; @Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. In reality, bacteria and phage communities are dynamic and complex. They frequently share genetic information and work together to maintain stable ecosystems. Removal of bacteria or phage can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. Studying such bacteria-phage interactions through community-wide networks built from inferred relationships could, e.g., offer insights into the drivers of human microbiome diversity across body sites and enable the study of human microbiome network dynamics overall. <!--This is not the right place to put a concluding clause (it is both illogical paragraph/idea flow, but also redundant to the closing sentence of the last paragraph...where concluding clauses fit well), I tried to morph it into more forward looking wording ("what do we need?"). Then the next paragraph starts with "and here is what we provided towards what we need." -->

In this study, human-associated bacterial and phage communities were characterized by their inferred relationships using three published paired virus and bacteria-dominated whole community metagenomic datasets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Machine learning and graph theory techniques were leveraged to establish and explore the human bacteria-phage network diversity therein. This approach built upon previous large-scale phage-bacteria network analyses by inferring interactions from metagenomic datasets, rather than culture-dependent data [@Flores:2013hc], which is limited in the scale of possible experiments and analyses. Our metagenomic interaction inference model improves upon previous models of phage-host predictions that have utilized a variety of techniques, such as linear models to predict bacteria-phage co-occurence based on taxonomy assignments <!-- does this capture your understanding of their approach?--> [@LimaMendez:2015hw], and nucleotide similarity models applied to both whole virus genomes [@Edwards:2015iz] and related clusters of whole and partial virus genomes [@@Roux:2016cc]<!--neither CRISPRs nor OGUs had been introduced yet, so I simplified and took those out here. Otherwise, this is the type of overview I'd envisioned, thanks...there is no need to mention that these are in ocean systems, only focus on giving credit to the approaches and methods.-->. Our approach uniquely included protein interaction data and was validated based on experimentally determined positive and negative interactions, i.e., "who does and does not infect whom." Through this approach, we were able to provide a basic understanding of the network dynamics associated with phage and bacterial communities on and in the human body. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support ecological communities with distinct communication structures and propensities toward community instability. Through an improved understanding of network structures across the human body, we empower future studies to investigate how these communities and interaction dynamics are influenced by disease state and the overall impact they may have on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in virus-bacteria interaction networks across healthy human bodies by leveraging previously published shotgun sequence datasets of purified viral metagenomes (viromes) paired with bacteria-dominated whole community metagenomes. Our study contained three datasets that explored the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. <!--MDA reference not needed/relevant here, distracting-->These datasets were selected because their virome samples were subjected to virus-like particle (VLP) purification. To this end, they employed combinations of filtration, chloroform/DNase treatment, and cesium chloride gradients to eliminate organismal DNA and thereby allow for direct assessment of both the extracellular and fully-assembled intracellular virome **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. While the whole metagenomic shotgun sequence samples were not subjected to purification, they primarily consisted of bacteria [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei].

@@

Four phage and bacterial genomic features were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) bacterial Clustered Regularly Interspaced Short Palindromic Repeat (CRISPR) spacer sequences that target phages, and 4) similarity of protein families associated with experimentally identified protein-protein interactions [@Orchard:2014hq]. The resulting random forest model was assessed and the area under its receiver operating characteristic (ROC) curve was `r signif(genmodelper$ROC, digits = sigfig)`, the model sensitivity was `r signif(genmodelper$Sens, digits = sigfig)`, and specificity was `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was amino acid similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, and CRISPRs were largely uninformative, due to the lack of identifiable CRISPRs in the dataset and their redundancy with the nucleotide similarity methods **(Figure \ref{RocCurve} B)**. Approximately one third of the training set relationships yielded no score and therefore were unable to be assigned an interaction prediction **(Figure \ref{RocCurve} C)**.

We used the random forest model to classify the relationships between bacteria and phage operational genomic units, which were then used to build the interactive network. The master network contained the three studies as sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and OGU abundance within the community were stored in the master network for downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, representing `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` bacteria. Although the network was highly connected, not all relationships were present in all samples. As relationships were weighted by the relative abundances of their associated bacteria and phages, lowly abundant relationships could be present but not highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Vertices"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], digits = 1)`% and `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Edges"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], digits = 1)`% of the master network nodes and edges, respectively **(Figure \ref{RocCurve} E - F)**. The phages and bacteria in the gut diet and twin sample sets were more sparsely related: each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")` **(Figure \ref{RocCurve} E - F)**.

@@

The average centrality of each sample was quantified using the weighted eigenvector centrality metric. Intermittently moist skin sites (dynamic sites that fluctuate between being moist and dry) were significantly less connected than the more stable moist and sebaceous environments (p-value < 0.001, **Figure \ref{skinnetwork} A)**. Also, skin sites that were protected from the environment (occluded) were much more highly connected than those that were constantly exposed to the environment or only intermittently occluded (p-value < 0.001, **Figure \ref{skinnetwork} B)**.

@@

# Discussion
Foundational work has provided a baseline understanding of the human microbiome by characterizing bacterial and viral diversity across the human body [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. Here, we offer an approach to describe how phage-bacteria networks differ throughout the human body to provide a baseline for future studies of how and why microbiome networks differ in disease states. We developed and implemented a network-based analytical model to evaluate the basic properties of the human microbiome through bacteria and phage relationships, instead of membership or diversity alone. This enabled the application of network theory to provide a new perspective on complex ecological communities. We utilized metrics of connectivity to model the extent to which communities of bacteria and phages interact through mechanisms such as horizontal gene transfer, modulated bacterial gene expression, and alterations in abundance.

Just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggested that the community properties inferred from microbiome interaction network structures, such as stability, the potential for horizontal gene transfer between members, co-evolution of populations, etc., were person-specific. These properties may be impacted by personal factors ranging from the body's immune system to external environmental conditions, such as climate and diet.

@@

Finally, the networks in this study were built using operational genomic units (OGUs), which represented groups of highly similar bacteria or phage genomes or genome fragments as clustered sub-populations. Similar clustering definition and validation methods, both computational and experimental, have been implemented in other metagenomic sequencing studies, as well [@Minot:2012ed; @Deng:2014eb; @Brum:2015iaa; @Roux:2016cc]. These approaches could offer yet another level of sophistication to our network-based analyses. While this operationally defined clustering approach allows us to study whole community networks, our ability to make conclusions about interactions among specific phage or bacterial species or populations is inherently limited. Future work must address this limitation, e.g., through improved binning methods and deeper metagenomic shotgun sequencing, but most importantly through an improved conceptual framing of what defines ecologically and evolutionarily cohesive units for both phage and bacteria, sensu Polz et al., 2006 [@Polz:2006fi]. Defining operational genomic units and their taxonomic underpinnings (e.g., whether OGU clusters represent genera or species) is an active area of work critical to the utility of this approach. As a first step, phylogenomic analyses have been performed to cluster cyanophage isolate genomes into informative groups using shared gene content, average nucleotide identity of shared genes, and pairwise differences between genomes [@Gregory:2016cg]. Such population-genetic assessment of phage evolution, coupled with the ecological implications of genome heterogeneity, will inform how to define nodes in future iterations of the ecological network developed here.

@@

https://github.com/SchlossLab/Hannigan_ConjunctisViribus_GenRes_2017

@@

##### Detect Matching Prophages within Bacterial Genomes
Temperate bacteriophages infect and integrate into their bacterial host's genome. We detected integrated phage elements within bacterial genomes by aligning phage genomes to bacterial genomes using the nucleotide BLAST algorithm and a minimum e-value of 1e-10. The resulting bitscore of each alignment was recorded for use in our classification model. <!-- I get suspicious whenever e-value is used as a threshold. Do you have anecdotal evidence this a is a good criterion? Should be easy to check-->

@@

# Acknowledgments
We thank the members of the Schloss lab for their underlying contributions. GDH was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528). GDH and PDS were supported in part by funding from the NIH (P30DK034933, U19AI09087, and U01AI124255).

>>>>

***Abstract Length***: 236 / 250

***Importance Length***: 145 / 150

***Text Length***: 4,987 / 5,000 Words

@@

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Most previous work has relied on studying microbes and viruses independently, thereby reducing them to two separate communities. Such approaches are unable to capture how these microbial communities interact, such as through processes that maintain community stability or allow phage-host populations to co-evolve. We developed and implemented a network-based analytical approach to describe phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages could infect which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate phage-bacteria connectedness across the human body. We found that gut and skin network structures were person-specific and not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. Network structure differed between skin sites, with those exposed to the external environment being less connected and more prone to instability. This study quantified and contrasted the diversity of virome-microbiome networks across the human body and illustrated how environmental factors may influence phage-bacteria interactive dynamics. This work provides a baseline for future studies to better understand system perturbations, such as disease states, through ecological networks.

# Importance

The human microbiome, the collection of microbial communities that colonize the human body, is a crucial component to health and disease. Two major components to the human microbiome are the bacterial and viral communities. These communities have primarily been studied separately using metrics of community composition and diversity. These approaches have failed to capture the complex dynamics of interacting bacteria and phage communities, which frequently share genetic information and work together to maintain stable ecosystems. Removal of bacteria or phage can disrupt or even collapse those ecosystems. Relationship-based network approaches allow us to capture this interaction information. Using this network-based approach with three independent human cohorts, we were able to present an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how and why microbiome networks differ in disease states.

@@

Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Bacterial communities have been associated with disease states, including a range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases, such as inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz] and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered human viromes (virus communities consisting primarily of bacteriophages) also have been  associated with diseases and perturbations, including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], spread of antibiotic resistance [@Modi:2013fia], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. Viruses act in concert with their microbial hosts as a single ecological community [@Haerter:2014ii]. Viruses influence their living microbial host communities through processes including lysis, host gene expression modulation [@Lindell:2005gz, @Tyler:2013fl, @Hargreaves:2014ja], influence on evolutionary processes such as horizontal gene transfer [@Moon:2015fa, @Modi:2013fi, @Ogg:1981th, @Frost:2005dn] or antagonistic co-evolution [@Koskella:2014ds], and alteration of ecosystem processes and elemental stoichiometry [@Jover:2014gq].

Previous human microbiome work has focused on bacterial and viral communities, but have reduced them to two separate communities by studying them independently  [@Norman:2015kb; @Manrique:2016dx; @Ly:2014ew; @Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. This approach fails to capture the complex dynamics of interacting bacteria and phage communities, which frequently share genetic information and work together to maintain stable ecosystems. Removal of bacteria or phage can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. Relationship-based network approaches allow us to capture this interaction information. Studying such bacteria-phage interactions through community-wide networks built from inferred relationships could offer further insights into the drivers of human microbiome diversity across body sites and enable the study of human microbiome network dynamics overall.

In this study, we characterized human-associated bacterial and phage communities by their inferred relationships using three published paired virus and bacteria-dominated whole community metagenomic datasets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei].  We leveraged machine learning and graph theory techniques to establish and explore the human bacteria-phage network diversity therein. This approach built upon previous large-scale phage-bacteria network analyses by inferring interactions from metagenomic datasets, rather than culture-dependent data [@Flores:2013hc], which is limited in the scale of possible experiments and analyses. Our metagenomic interaction inference model improved upon previous models of phage-host predictions that have utilized a variety of techniques, such as linear models to predict bacteria-phage co-occurrence using taxonomic assignments [@LimaMendez:2015hw], and nucleotide similarity models that were applied to both whole virus genomes [@Edwards:2015iz] and related clusters of whole and partial virus genomes [@@Roux:2016cc]. Our approach uniquely included protein interaction data and was validated based on experimentally determined positive and negative interactions (i.e. who does and does not infect whom). Through this approach we were able to provide a basic understanding of the network dynamics associated with phage and bacterial communities on and in the human body. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support ecological communities with distinct communication structures and propensities toward community instability. Through an improved understanding of network structures across the human body, we empower future studies to investigate how these communities dynamics are influenced by disease states and the overall impact they may have on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in virus-bacteria interaction networks across healthy human bodies by leveraging previously published shotgun sequence datasets of purified viral metagenomes (viromes) paired with bacteria-dominated whole community metagenomes. Our study contained three datasets that explored the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. We selected these datasets because their virome samples were subjected to virus-like particle (VLP) purification. To this end, they employed combinations of filtration, chloroform/DNase treatment, and cesium chloride gradients to eliminate organismal DNA and thereby allow for direct assessment of both the extracellular and fully-assembled intracellular virome **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. While the whole metagenomic shotgun sequence samples were not subjected to purification, they primarily consisted of bacteria [@Minot:2011ez; @Hannigan:2015fz; @Reyes:2010cwa; @Turnbaugh:2009ei].

@@

Four phage and bacterial genomic features were used in a random forest model to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) bacterial Clustered Regularly Interspaced Short Palindromic Repeat (CRISPR) spacer sequences that target phages, and 4) similarity of protein families associated with experimentally identified protein-protein interactions [@Orchard:2014hq]. The resulting random forest model was assessed and the area under its receiver operating characteristic (ROC) curve was `r signif(genmodelper$ROC, digits = sigfig)`, the model sensitivity was `r signif(genmodelper$Sens, digits = sigfig)`, and specificity was `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was amino acid similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, and CRISPRs were largely uninformative, due to the minimal amount of identifiable CRISPRs in the dataset and their redundancy with the nucleotide similarity methods **(Figure \ref{RocCurve} B)**. Approximately one third of the training set relationships yielded no score and therefore were unable to be assigned an interaction prediction **(Figure \ref{RocCurve} C)**.

We used our random forest model to classify the relationships between bacteria and phage operational genomic units, which were then used to build the interactive network. The master network contained the three studies as sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and OGU abundance within the community were stored in the master network for parsing in downstream analyses **(Supplemental Figure \ref{NetworkDiagram})**. The master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, representing `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` bacteria. Although the network was highly connected, not all relationships were present in all samples. As relationships were weighted by the relative abundances of their associated bacteria and phages, lowly abundant relationships could be present but not highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Vertices"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], digits = 1)`% and `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Edges"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], digits = 1)`% of the master network nodes and edges, respectively **(Figure \ref{RocCurve} E - F)**. The phages and bacteria in the gut diet and twin sample sets were more sparsely related: each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")` **(Figure \ref{RocCurve} E - F)**.

@@

The average centrality of each sample was quantified using the weighted eigenvector centrality metric. Intermittently moist skin sites (dynamic sites that fluctuate between being moist and dry) were significantly less connected than the more stable moist and sebaceous environments (p-value < 0.001, **Figure \ref{skinnetwork} A)**. Also, skin sites that were occluded from the environment were much more highly connected than those that were constantly exposed to the environment or only intermittently occluded (p-value < 0.001, **Figure \ref{skinnetwork} B)**.

@@

# Discussion
Foundational work has provided a baseline understanding of the human microbiome by characterizing bacterial and viral diversity across the human body [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. Here, we offer an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how and why microbiome networks differ in disease states. We developed and implemented a network-based analytical model to evaluate the basic properties of the human microbiome through bacteria and phage relationships, instead of membership or diversity alone. This enabled the application of network theory to provide a new perspective on complex ecological communities. We utilized metrics of connectivity to model the extent to which communities of bacteria and phages interact through mechanisms such as horizontal gene transfer, modulated bacterial gene expression, and alterations in abundance.

Just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggested that the community properties inferred from microbiome interaction network structures, such as stability, the potential for horizontal gene transfer between members, and co-evolution of populations, were person-specific. These properties may be impacted by personal factors ranging from the body's immune system to external environmental conditions, such as climate and diet.

@@

Finally, the networks in this study were built using operational genomic units (OGUs), which represented groups of highly similar bacteria or phage genomes or genome fragments as clustered sub-populations. Similar clustering definition and validation methods, both computational and experimental, have been implemented in other metagenomic sequencing studies, as well [@Minot:2012ed; @Deng:2014eb; @Brum:2015iaa; @Roux:2016cc]. These approaches could offer yet another level of sophistication to our network-based analyses. While this operationally defined clustering approach allows us to study whole community networks, our ability to make conclusions about interactions among specific phage or bacterial species or populations is inherently limited. Future work must address this limitation, e.g., through improved binning methods and deeper metagenomic shotgun sequencing, but most importantly through an improved conceptual framing of what defines ecologically and evolutionarily cohesive units for both phage and bacteria [@Polz:2006fi]. Defining operational genomic units and their taxonomic underpinnings (e.g., whether OGU clusters represent genera or species) is an active area of work critical to the utility of this approach. As a first step, phylogenomic analyses have been performed to cluster cyanophage isolate genomes into informative groups using shared gene content, average nucleotide identity of shared genes, and pairwise differences between genomes [@Gregory:2016cg]. Such population-genetic assessment of phage evolution, coupled with the ecological implications of genome heterogeneity, will inform how to define nodes in future iterations of the ecological network developed here.

@@

https://github.com/SchlossLab/Hannigan_ConjunctisViribus_mSystems_2017

@@

##### Detect Matching Prophages within Bacterial Genomes
Temperate bacteriophages infect and integrate into their bacterial host's genome. We detected integrated phage elements within bacterial genomes by aligning phage genomes to bacterial genomes using the nucleotide BLAST algorithm and a minimum e-value of 1e-10. The resulting bitscore of each alignment was recorded for use in our classification model.

@@

# Acknowledgments
We thank the members of the Schloss lab for their underlying contributions.

# Funding Information
GDH was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528). GDH and PDS were supported in part by funding from the NIH (P30DK034933, U19AI09087, and U01AI124255).
